prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-25 inst/MyArticle/paper.Rmd

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
      citation_package: natbib
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of signal processing software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easy. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis application development easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

Several data analysis packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` determines the statistical model from the structure of the sample annotation, allowing users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows the specification of a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` combines two models: robust linear models fitted to protein abundances and a quasibinomial generalized linear model fitted to peptide counts, into Hurdle model. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

Of note are the various approaches to handling missing observations, which are inherent in mass spectrometric bottom-up experiments. For instance, `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and imputation. Other means of modeling missing observations are the Hurdle model discussed by @goeminne2020msqrob, while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

@@

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables us to call various machine learning (ML) methods, and makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the differential expression analysis of quantitative proteomics data. However, the existing packages differ widely regarding supported designs, model specifications, and output formats. At the same time, they have in common: they fit linear models to either peptide or protein intensities, determine differences among groups, and afterward apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented (OO) design, with R linear models as a core, and add features such as $p$-value aggregation, variance shrinkage, or modeling of missing observations.

Furthermore, the functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish  [@tukey1977exploratory], robust linear models [@goeminne2020msqrob]. Furthermore, peptide or protein abundances can be scaled and transformed using robust scaling, _quantile_ normalization or `r BiocStyle::Biocpkg("vsn")` to remove systematic differences among samples and heteroscedasticity. In this respect, `r BiocStyle::Githubpkg("fgcz/prolfqua")` can be compared with R packages such as `r BiocStyle::Biocpkg("DEP")` [@DEP2018] or `r BiocStyle::Biocpkg("POMA")` [@POMA2021] which support the entire differential expression analysis pipeline.

We also implemented functionality and use the Ionstar [@shen2018ionstar] and CPTAC (available from `r BiocStyle::Biocpkg("msdata")`) dataset to benchmark the modeling methods implemented within `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$).

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r BiocStyle::Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences in column names among the various input formats and enables the use of the methods without additional arguments. We show an example code for creating an MSFragger[@yu2020fast] configuration in the Appendix. For popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy data. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (see Supplementary Figure SX).

>>>>

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
      citation_package: natbib
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of signal processing software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easily. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis application development easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

Several data analysis packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` determines the statistical model from the structure of the sample annotation, which allows users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows the specification of a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r BiocStyle::Biocpkg("PECA")` performs a roll-up of peptide level differences and peptide level $p$-value estimates obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` combines two models: robust linear models fitted to protein abundances and a quasibinomial generalized linear model fitted to peptide counts, into Hurdle model. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation.

Of note are the various approaches to handle missing observations, which are inherent in mass spectrometric bottom-up experiments. For instance, `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and imputation. Other means of modeling missing observations are the Hurdle models discussed by @goeminne2020msqrob, while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

@@

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008] which enables us to call various Machine Learning (ML) methods, and makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the differential expression analysis of quantitative proteomics data. However, the existing packages differ widely regarding supported designs, model specifications, and output formats. At the same time, they have the following features in common: fitting linear models to either peptide or protein intensities, determining differences among groups, and afterward applying empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented (OO) design, with R linear models as a core, and add features such as $p$-value aggregation, variance shrinkage, or modeling of missing observations.

Furthermore, `r BiocStyle::Githubpkg("fgcz/prolfqua")` also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N [@Grossmann2010], Tukey's median polish  [@tukey1977exploratory], robust linear models [@goeminne2020msqrob]. Peptide or protein abundances can then be scaled and transformed using robust scaling, _quantile_ normalization or `r BiocStyle::Biocpkg("vsn")` to remove systematic differences among samples and heteroscedasticity. In this respect, `r BiocStyle::Githubpkg("fgcz/prolfqua")` can be compared with R packages such as `r BiocStyle::Biocpkg("DEP")` [@DEP2018] or `r BiocStyle::Biocpkg("POMA")` [@POMA2021] which support the entire differential expression analysis pipeline.

We use the Ionstar [@shen2018ionstar] and CPTAC (available from `r BiocStyle::Biocpkg("msdata")`) dataset to benchmark the modeling methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` and to compare our results with those of `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`. Since group sizes are relatively small, typically with four or five subjects per group, the high power of the tests is a relevant criterion to assess the modeling method. The quantified proteins can be ranked using the estimated fold-change, $t$-statistics, or scaled $p$-value, and afterward subjected to gene set enrichment (GSEA) or over-representation analysis [@subramanian2005gene] to determine up or down-regulated groups of proteins. Furthermore, the statistical model must provide an unbiased estimate of the false discovery rate (FDR) to manage expectations when selecting protein lists for follow-up experiments. We will use the partial area under the receiver operator curve (ROC) to assess the power of the tests and compare the FDR with the false discovery proportion ($FDP$).

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r BiocStyle::Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences among the various input formats in column names and enables the use of the methods without additional arguments. An example code for creating an MSFragger[@yu2020fast] configuration can be found in the Appendix. We implemented methods that transform the data into tidy data for popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (see Supplementary Figure SX).
